---
title: "Analysis_szg"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
```

### Some observations from the data

```{r}
weekly_deaths_data <- read_csv("data/weekly_deaths_by_state_and_causes.csv") %>% 
  select(-1)

# Get the data I need to analysis 
data_sg <- weekly_deaths_data %>%
  filter(`Jurisdiction of Occurrence` %in% c("Massachusetts", "Michigan", "Minnesota", 
                                             "Mississippi", "Missouri", "Montana", 
                                             "Nebraska", "Nevada", "New Hampshire", 
                                             "New Jersey")) %>% 
  select(1:19) %>% 
  rename(
    `Septicemia` = `Septicemia (A40-A41)`, 
    `Malignant neoplasms` = `Malignant neoplasms (C00-C97)`, 
    `Diabetes mellitus` = `Diabetes mellitus (E10-E14)`, 
    `Alzheimer disease` = `Alzheimer disease (G30)`, 
    `Influenza and pneumonia` = `Influenza and pneumonia (J09-J18)`, 
    `Chronic lower respiratory diseases` = `Chronic lower respiratory diseases (J40-J47)`, 
    `Other respiratory diseases` = `Other diseases of respiratory system (J00-J06,J30-J39,J67,J70-J98)`, 
    `Kidney Disease` = `Nephritis, nephrotic syndrome and nephrosis (N00-N07,N17-N19,N25-N27)`, 
    `Other Symptoms` = `Symptoms, signs and abnormal clinical and laboratory findings, not elsewhere classified (R00-R99)`, 
    `Diseases of heart` = `Diseases of heart (I00-I09,I11,I13,I20-I51)`, 
    `Cerebrovascular diseases` = `Cerebrovascular diseases (I60-I69)`, 
    `COVID-19 (Multiple)` = `COVID-19 (U071, Multiple Cause of Death)`, 
    `COVID-19 (Underlying)` = `COVID-19 (U071, Underlying Cause of Death)`
  )
state_data_list <- split(data_sg, data_sg %>% pull(`Jurisdiction of Occurrence`))
#summary(data_sg[ ,5:19])
```

Overall, the columns with fewer missing values are Malignant neoplasms, Diabetes mellitus, Alzheimer disease, Chronic lower respiratory diseases, Diseases of heart, Cerebrovascular diseases, COVID-19 (Multiple), and COVID-19 (Underlying).  

```{r, message = FALSE, results = "hide"}
data_ma <- state_data_list[["Massachusetts"]]
summary(data_ma[ ,5:19])

data_mi <- state_data_list[["Michigan"]]
summary(data_mi[ ,5:19])

data_mn <- state_data_list[["Minnesota"]]
summary(data_mn[ ,5:19])

data_ms <- state_data_list[["Mississippi"]]
summary(data_ms[ ,5:19])

data_mo <- state_data_list[["Missouri"]]
summary(data_mo[ ,5:19])

data_mt <- state_data_list[["Montana"]]
summary(data_mt[ ,5:19])

data_ne <- state_data_list[["Nebraska"]]
summary(data_ne[ ,5:19])

data_nv <- state_data_list[["Nevada"]]
summary(data_nv[ ,5:19])

data_nh <- state_data_list[["New Hampshire"]]
summary(data_nh[ ,5:19])

data_nj <- state_data_list[["New Jersey"]]
summary(data_nj[ ,5:19])
```

Distribution of NAs:  
*Massachusetts*: nice!  
*Minnesota*: Septicemia.  
*Mississippi*: Septicemia, Other respiratory diseases, Other Symptoms  
*Missouri*: Other Symptoms  
*Montana*: Septicemia, Diabetes mellitus, Alzheimer disease, Influenza and pneumonia, Other respiratory diseases, Kidney Disease, Other Symptoms, Cerebrovascular diseases.   
*Nebraska*: Septicemia, Influenza and pneumonia, Other respiratory diseases, Kidney Disease, Other Symptoms, Diseases of heart   
*Nevada*: Septicemia, Other respiratory diseases, Kidney Disease, Other Symptoms   
*New Hampshire*: Septicemia, Diabetes mellitus, Alzheimer disease, Influenza and pneumonia,Other respiratory diseases, Kidney Disease, Other Symptoms, COVID-19 (Underlying)   
*New Jersey*: nice!  

Montana, Nebraska and New Hampshire have a severe data missing problem. 

### Perform some time series analysis on selected state 

```{r, warning=FALSE}
# Prepare the data
data_ma_long <- data_ma %>%
  select(`Week Ending Date`,
         `Septicemia`, `Malignant neoplasms`, `Diabetes mellitus`, `Alzheimer disease`,
         `Influenza and pneumonia`, `Chronic lower respiratory diseases`, `Other respiratory diseases`,
         `Kidney Disease`, `Other Symptoms`,
         `Diseases of heart`, `Cerebrovascular diseases`,
         `COVID-19 (Multiple)`, `COVID-19 (Underlying)`) %>%
  pivot_longer(
    cols = -`Week Ending Date`,
    names_to = "Disease",
    values_to = "Deaths"
  )

# Create the interactive plot using plotly
plot_ma <- plot_ly(
  data_ma_long,
  x = ~`Week Ending Date`,
  y = ~Deaths,
  color = ~Disease,
  type = 'scatter',
  mode = 'lines'
) %>%
  layout(
    title = "Trends in Death Counts Over Time by Disease in MA",
    xaxis = list(title = "Week Ending Date"),
    yaxis = list(title = "Number of Deaths"),
    legend = list(title = list(text = "Disease"))
  )

# Display the plot
plot_ma
```