---
title: "Analysis_yhc"
output: html_document
date: "2024-11-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(lubridate)
library(tidyr)

```
## Import and clean data
```{r}
data=read_csv("./data/weekly_deaths_by_state_and_causes.csv") 
states <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", 
            "Colorado", "Connecticut", "Delaware", "Florida", "Georgia", 
            "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", 
            "Kentucky", "Louisiana", "Maine", "Maryland")

yh_data <- data |>
  janitor::clean_names() |>
  rename_with(~ str_replace_all(., " ", "_")) |>
  filter(jurisdiction_of_occurrence %in% states) |>
  rename_with(~ make.unique(str_replace(., "_\\w\\d.*", ""))) |>
  mutate(
    across(where(is.character), ~ ifelse(. == "Suppressed (counts 1-9)", 
                                         as.character(sample(1:9, 1)), .))
  )  |>
  mutate(month = month(week_ending_date)) 
```
## Visualize the extent of missing data
```{r fig.width=18, fig.height=10}
data_missing = yh_data |>
  pivot_longer(cols = 6:20, names_to = "disease", values_to = "value") |>
  mutate(missing = ifelse(is.na(value), 1, 0))  |> # 1 as missing
  group_by(jurisdiction_of_occurrence, disease) |>
  summarize(missing_rate = mean(is.na(value)), .groups = "drop")

missing_plot=
ggplot(data_missing, aes(x = jurisdiction_of_occurrence, y = disease, fill = missing_rate)) +
  geom_tile(color = "white") +  
  scale_fill_gradient(
    low = "white",    # no missing as white
    high = "red",     # all missing as red
    na.value = "gray" 
  ) +
  labs(
    title = "Missing Data Heatmap",
    x = "State",
    y = "Disease",
    fill = "Missing Rate"
  ) +theme(
    axis.text.y = element_text(size = 10, angle = 45),
    plot.title = element_text(hjust = 0.5)
  )



missing_plot
```
It can be observed that the states with relatively complete data are Alabama, Arizona, California, Florida, Georgia, Illinois, Indiana, Kentucky, and Louisiana.
##Week with the Highest Mortality in the Past Four Years
```{r fig.width=18, fig.height=10}
total_weekly_deaths = yh_data |>
  group_by(mmwr_week) |>
  summarize(total_deaths = sum(natural_cause, na.rm = TRUE)) |>
  ungroup()

total_weekly_deaths_plot=
ggplot(total_weekly_deaths, aes(x = factor(mmwr_week), 
                                y = total_deaths, fill = factor(mmwr_week))) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Total Mortality for Each Week Over the Past Four Years",
    x = "Week",
    y = "Total Death Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

total_weekly_deaths_plot
```
##Mortality Trends Over Time for Different Causes of Death
```{r fig.width=18, fig.height=10}
mortality_trends_data = yh_data |>
  select(
    week_ending_date, 
    jurisdiction_of_occurrence, 
    mmwr_year,
    septicemia,
    malignant_neoplasms,
    diabetes_mellitus,
    alzheimer_disease,
    influenza_and_pneumonia,
    chronic_lower_respiratory_diseases,
    other_diseases_of_respiratory_system,
    nephritis_nephrotic_syndrome_and_nephrosis,
    symptoms_signs_and_abnormal_clinical_and_laboratory_findings_not_elsewhere_classified,
    diseases_of_heart,
    cerebrovascular_diseases,
    covid,
    covid.1
  ) |>
  gather(key = "cause_of_death", value = "death_count", 
         septicemia,
         malignant_neoplasms,
         diabetes_mellitus,
         alzheimer_disease,
         influenza_and_pneumonia,
         chronic_lower_respiratory_diseases,
         other_diseases_of_respiratory_system,
         nephritis_nephrotic_syndrome_and_nephrosis,
         symptoms_signs_and_abnormal_clinical_and_laboratory_findings_not_elsewhere_classified,
         diseases_of_heart,
         cerebrovascular_diseases,
         covid,
         covid.1) |>
  group_by(week_ending_date, cause_of_death) |>
  summarize(total_deaths = sum(death_count, na.rm = TRUE)) |>
  ungroup()

mortality_trends_plot=
ggplot(mortality_trends_data, aes(x = week_ending_date, 
                                  y = total_deaths,
                                  color = cause_of_death, 
                                  group = cause_of_death)) +
  geom_line(alpha=0.6) +
  geom_point(alpha=0.6) +
  theme_minimal() +
  labs(
    title = "Mortality Trends Over Time for Different Causes of Death",
    x = "Week Ending Date",
    y = "Total Deaths",
    color = "Cause of Death"
  ) +
theme(
    plot.title = element_text(hjust = 0.5)
  )
mortality_trends_plot
```
##Changes in the Proportion of Each Cause of Death in Total Mortality Over Four Years
```{r fig.width=18, fig.height=10}
mortality_data <- yh_data |>
  select(
    week_ending_date, 
    jurisdiction_of_occurrence, 
    mmwr_year,
    septicemia,
    malignant_neoplasms,
    diabetes_mellitus,
    alzheimer_disease,
    influenza_and_pneumonia,
    chronic_lower_respiratory_diseases,
    other_diseases_of_respiratory_system,
    nephritis_nephrotic_syndrome_and_nephrosis,
    symptoms_signs_and_abnormal_clinical_and_laboratory_findings_not_elsewhere_classified,
    diseases_of_heart,
    cerebrovascular_diseases,
    covid,
    covid.1
  ) |>
  gather(key = "cause_of_death", value = "death_count", 
         septicemia,
         malignant_neoplasms,
         diabetes_mellitus,
         alzheimer_disease,
         influenza_and_pneumonia,
         chronic_lower_respiratory_diseases,
         other_diseases_of_respiratory_system,
         nephritis_nephrotic_syndrome_and_nephrosis,
         symptoms_signs_and_abnormal_clinical_and_laboratory_findings_not_elsewhere_classified,
         diseases_of_heart,
         cerebrovascular_diseases,
         covid,
         covid.1) |>
  group_by(week_ending_date, mmwr_year, cause_of_death) |>
  summarize(death_count = sum(death_count, na.rm = TRUE)) |>
  group_by(week_ending_date, mmwr_year) |>
  mutate(total_deaths = sum(death_count, na.rm = TRUE)) |>
  mutate(proportion = death_count / total_deaths) |>
  ungroup()

mortality_plot=
ggplot(mortality_data, aes(x = week_ending_date, y = proportion, fill = cause_of_death)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Changes in the Proportion of Each Cause of Death in Total Mortality Over Four Years",
    x = "Week Ending Date",
    y = "Proportion of Total Deaths",
    fill = "Cause of Death"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")+
  theme(
    plot.title = element_text(hjust = 0.5)
  )

mortality_plot


```

