---
title: "analysis_az"
author: "Alice Zhou"
date: "2024-11-18"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---


# EDA
## data overview

The original dataset (10,476 rows and 35 columns) contains weekly counts of death by cause, week and geographical location.

Date Range: Weekly data from 2020 to 2023.

Geographic Scope: Includes state-level data and "United States" as an aggregated entry.


### Key Variables:
* Date Variables: Data As Of, Week Ending Date, MMWR Year, MMWR Week.
* Mortality Data: Includes counts for "All Cause", "Natural Cause", and specific causes like:
  * Septicemia (A40-A41)
  * Malignant neoplasms (C00-C97)
  * Diabetes mellitus (E10-E14)
  * Alzheimer disease (G30)
  * Influenza and pneumonia (J09-J18)
  * Chronic lower respiratory diseases (J40-J47)
  * COVID-19 (U071, Multiple Cause of Death) and COVID-19 (U071, Underlying Cause of Death).
  * Flags: Columns starting with flag_indicate potential data quality issues for specific cause counts.

### Observations:
Due to privacy protection, cells with count under 10 has been supressed and shown as blank. We will generate a random whole number between 1-10 for those cells for further analyses.

```{r setup, message = FALSE}
library(tidyverse)
library(lubridate)
library(viridis)
library(broom)
library(plotly)
library(datasets)
library(usmap)
```

```{r}
set.seed(123)
```

Import data and do some inital cleaning
```{r}
data = read_csv("data/weekly_deaths_by_state_and_causes.csv") %>% 
  rename(
    `Septicemia` = `Septicemia (A40-A41)`, 
    `Malignant neoplasms` = `Malignant neoplasms (C00-C97)`, 
    `Diabetes mellitus` = `Diabetes mellitus (E10-E14)`, 
    `Alzheimer disease` = `Alzheimer disease (G30)`, 
    `Influenza and pneumonia` = `Influenza and pneumonia (J09-J18)`, 
    `Chronic lower respiratory diseases` = `Chronic lower respiratory diseases (J40-J47)`, 
    `Other respiratory diseases` = `Other diseases of respiratory system (J00-J06,J30-J39,J67,J70-J98)`, 
    `Kidney Disease` = `Nephritis, nephrotic syndrome and nephrosis (N00-N07,N17-N19,N25-N27)`, 
    `Other Symptoms` = `Symptoms, signs and abnormal clinical and laboratory findings, not elsewhere classified (R00-R99)`, 
    `Diseases of heart` = `Diseases of heart (I00-I09,I11,I13,I20-I51)`, 
    `Cerebrovascular diseases` = `Cerebrovascular diseases (I60-I69)`, 
    `COVID-19 (Multiple)` = `COVID-19 (U071, Multiple Cause of Death)`, 
    `COVID-19 (Underlying)` = `COVID-19 (U071, Underlying Cause of Death)`
  ) %>% 
  janitor::clean_names() %>% 
  rename(state = jurisdiction_of_occurrence) %>% 
  mutate(data_as_of = mdy(`data_as_of`),
         year = year(week_ending_date),
         month = month(week_ending_date)
         ) %>%  
  select(state, year, month, mmwr_week, week_ending_date, all_cause: covid_19_underlying)

data
```
for supressed cells, replace with a random number between 1-9
```{r}
# generate random number 1-9 to suppressed cells
data[] = lapply(data, function(col) {
  if (is.numeric(col)) { 
    col[is.na(col)] <- sample(1:9, sum(is.na(col)), replace = TRUE)
  }
  return(col)
})

summary(data)

write_csv(data, "data/full_weekly_deaths_by_state_and_causes.csv")
```

```{r}
subset_df = data %>%
  filter(state %in% c( "South Dakota",
 "Tennessee",
 "Texas",
 "Utah",
 "Vermont",
 "Virginia",
 "Washington",
 "West Virginia",
 "Wisconsin",
 "Wyoming"))

#inspect missing values state by state
count_na = function(df, state_name) {
  out_df = df %>% 
  filter(state == state_name) %>% 
  summarize(across(everything(), ~ sum(is.na(.)))) %>% 
  select(-state, -year, -month, -mmwr_week) %>% 
  pivot_longer(
    cols = everything(),
    names_to = "disease",
    values_to = "na"
  ) %>% 
    filter(na != 0)
  return(out_df)

}

count_na(subset_df,"South Dakota")
count_na(subset_df,"Tennessee")
count_na(subset_df,"Texas")
count_na(subset_df,"Utah")
count_na(subset_df,"Vermont")
count_na(subset_df,"Virginia")
count_na(subset_df,"Washington")
count_na(subset_df,"West Virginia")
count_na(subset_df,"Wisconsin")
count_na(subset_df,"Wyoming")

# significant missing was found in: South Dakota, Utah, Vermont, West Virginia, and Wyoming 

```
# Some general questions

-how many people die each year from 2020-2023 (what's the proportion of natural and non-natural causes?)
  -and, what's the mortality rate of the general population?

-which disease among the listed creates the most health burdens, that is, gives rise to the most deaths?

-correlation between diseases (heatmap?)

# Time Seires Analyses
How many people died from covid in 2020, 2021, 2022, and 2023?
```{r}

# Summarize COVID-19 deaths by year (both multiple and underlying cause)
covid_deaths_by_year = subset_df %>%
  group_by(year) %>%
  summarise(
    covid_deaths_multiple = sum(covid_19_multiple, na.rm = TRUE),
    covid_deaths_underlying = sum(covid_19_underlying, na.rm = TRUE)
  ) %>%
  filter(year %in% c(2020, 2021, 2022, 2023))

# Reshape the data for easier plotting
covid_deaths_long = covid_deaths_by_year %>%
  pivot_longer(cols = c(covid_deaths_multiple, covid_deaths_underlying),
               names_to = "cause_type", values_to = "deaths")

# Create the bar plot
covid_bar_plot = ggplot(covid_deaths_long, aes(x = factor(year), y = deaths, fill = cause_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "COVID-19 Deaths by Year (2020-2023)",
       x = "Year", y = "Number of Deaths", fill = "Cause Type") +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal()

ggplotly(covid_bar_plot)
```

Which months have the highest mortality?

How is deaths from specific diseases distribute across months? (which season will exacerbate what diseases?)

How often was COVID-19 the primary versus a secondary contributor to mortality?

# Correlational Analyses
Does COVID exacerbate other diseases?

Does specific conditions/diseases exacerbate covid?

Is there any correlation between cardiovascular and respiratory diseases?

Did the pandemic period affect mortality rates of non-COVID causes like heart disease or diabetes?

# geographical variations
Which states had higher mortality rates from specific causes, such as COVID-19 or heart disease, and did these vary over time?

________
# Explore the changes in COVID-19 deaths (underlying and multiple) over time.  

How did Covid-19 deaths in the US change over time (as underlying and multiple cause)?
```{r}
# Prepare the dataset
data_covid_us = data %>%
  filter(state == "United States") %>% 
  select(week_ending_date, covid_19_multiple, covid_19_underlying) %>%
  pivot_longer(cols = c(covid_19_multiple, covid_19_underlying),
               names_to = "cause_type",
               values_to = "deaths")

# Create an interactive line chart
plot = plot_ly(
  data = data_covid_us,
  x = ~week_ending_date,
  y = ~deaths,
  color = ~cause_type,
  colors = c("#440154", "#21908C"),
  type = 'scatter',
  mode = 'lines'
) %>%
  layout(
    title = "COVID-19 Deaths Over Time (Underlying vs. Multiple Cause)",
    xaxis = list(title = "Week Ending Date"),
    yaxis = list(title = "Number of Deaths"),
     legend = list(
      title = list(text = "Cause Type"), # Title for the legend
      orientation = "h",                # Horizontal legend
      x = 0.5,                          # Center the legend horizontally
      xanchor = "center",               # Set x-anchor to the center
      y = -0.2                           # Position the legend above the plot
    )
  )

# Show the plot
plot


```
## Key Findings
The plot yshows the progression of COVID-19 deaths in the United States over time, split between underlying and multiple cause deaths.

1. Initial Surge (March 2020 - May 2020)

There’s a sharp increase in both underlying and multiple cause deaths starting in March 2020, which corresponds to the initial outbreak and the declaration of the pandemic. This was likely when COVID-19 began to spread rapidly across the U.S.

The data shows a steep rise in deaths, especially in the multiple cause category, which may indicate that people were dying from COVID-19 along with other pre-existing conditions or complications.

2. Second Peak (Winter 2020 - Early 2021)

A second peak emerges in winter 2020 (Nov-Dec 2020), followed by a sustained high in early 2021. This aligns with the winter surge and the emergence of the Delta variant.

3. Decline After Vaccine Rollout (Mid 2021)

After mid-2021, we see a gradual decline in deaths as vaccines became more widely available and the country adapted to the pandemic. 

4. Fluctuations in 2022-2023

In 2022 and 2023, there are smaller peaks and valleys in the death counts, indicating continued fluctuations in the number of deaths due to COVID-19 as variants like Omicron and others continued to emerge.

The multiple cause deaths remain higher than underlying cause deaths in these later periods, suggesting that COVID-19 continued to be associated with other health conditions, especially in individuals with multiple risk factors.

5. Slow Decline into 2023

The plot shows a gradual reduction in both underlying and multiple cause deaths through 2023, but COVID-19 remains a significant cause of death for some time. This may reflect ongoing issues such as long COVID, healthcare access disparities, or specific high-risk populations.

Cases seems to accelerate during winter months.

```{r}
# Add a month column to the dataset
winter_deaths = data %>% 
  filter(state == "United States") %>% 
  mutate(month = month(week_ending_date, label = TRUE), # Extract month name
         season = case_when(                         # Define seasons
           month %in% c("Dec", "Jan", "Feb") ~ "Winter",
           TRUE ~ "Non-Winter"
         )) %>% 
  mutate(season = factor(season, levels = c("Winter", "Non-Winter")))

# compute a summary table
seasonal_summary <- winter_deaths %>%
  group_by(season) %>%
  summarize(
    total_underlying = sum(covid_19_underlying, na.rm = TRUE),
    total_multiple = sum(covid_19_multiple, na.rm = TRUE))


# Visualization: Boxplot
plot = plot_ly(
  data = winter_deaths,
  x = ~season,
  y = ~covid_19_underlying,
  type = "box",
  boxpoints = "all",  # Display individual data points
  jitter = 0.3,       # Add jitter to spread points
  pointpos = -1.8,    # Adjust position of points relative to the box
  marker = list(color = "blue", size = 5),
  line = list(color = "darkblue")
) %>%
  layout(
    title = "COVID-19 Underlying Deaths by Season",
    xaxis = list(title = "Season"),
    yaxis = list(title = "COVID-19 Underlying Deaths"),
    showlegend = FALSE
  )

# Display the plot
plot
seasonal_summary %>% knitr::kable()
```

```{r}
#check normality of outcome variable
qqnorm(winter_deaths$covid_19_underlying, main = "Q-Q Plot for COVID-19 Underlying Deaths")
qqline(winter_deaths$covid_19_underlying, col = "red", lwd = 2)
```
since the outcome variable is not normally distributed, we used a Wilcoxon Rank-Sum Test.

```{r}
wilcox.test(covid_19_underlying ~ season, data = winter_deaths) %>% 
  tidy() %>% 
  knitr::kable()
```
There is a significant difference between covid deaths in winter and non-winter months. We want to know whether this difference might be affected by other factors like a person's pre-existing conditions and co-morbidity (e.g., diabetes, heart disease, respiratory diseases).

# correlation between leading causes of death
```{r}
# Subset data for relevant columns
corr_mat_df = winter_deaths %>% 
  select(covid_19_underlying, septicemia, malignant_neoplasms, diabetes_mellitus, alzheimer_disease,influenza_and_pneumonia, chronic_lower_respiratory_diseases, other_respiratory_diseases, kidney_disease,
         diseases_of_heart, cerebrovascular_diseases) # Adjust based on your dataset

# Calculate correlation matrix
cor_matrix = cor(corr_mat_df, use = "complete.obs")

# Convert correlation matrix to long format
cor_long = as.data.frame(as.table(cor_matrix))

# Create heatmap
heatmap = plot_ly(
  data = cor_long,
  x = ~Var1,
  y = ~Var2,
  z = ~Freq,
  type = "heatmap",
  colorscale = "oranges"
) %>%
  layout(
    title = "Correlation Heatmap of COVID-19 and Other Diseases",
    xaxis = list(title = "Disease"),
    yaxis = list(title = "Disease"),
    colorbar = list(title = "Correlation")
  )

# Show heatmap
heatmap

```
In the time series analysis, we observed interesting pattern for diabetes, Alzheimer, and heart disease. The winter surge for these diseases seems to be lower in 2023 compared to 2021 and 2022. Combining with the heatmap, we hypothesize that the decline is due to those diseases' asscociation with COVID. To verify, we fit an MLR.

```{r}
# Fit a linear regression model
model = lm(
  covid_19_underlying ~ diabetes_mellitus + alzheimer_disease + diseases_of_heart,
  data = winter_deaths
)

summary(model) %>% tidy() %>% knitr::kable()
```

## How did covid death rate vary across states?
```{r}
# read the population data
population_data = read.table(file = "data/Population by States.txt", header = TRUE, sep = "\t", quote = "\"", fill = TRUE) %>% 
  as.tibble() %>% 
  janitor::clean_names() %>% 
  select(residence_state, year_code, population) %>% 
  rename(state = residence_state,
         year = year_code)

```
```{r}
# calculate annual covid deaths for each state
data_covid_states = 
  data %>% 
  filter(!state %in% c("United States", "Puerto Rico")) %>% 
  group_by(state, year) %>% 
  summarize(
    covid_deaths_total = sum(covid_19_multiple, na.rm = TRUE)
  )
```

join covid data with population data, calculate covid mortality rate
```{r}
data_covid_df = data_covid_states %>%
  left_join(population_data, by = c("year", "state")) %>% 
  mutate(
    covid_mortality = covid_deaths_total/population,
    covid_mortality_per100000 = covid_mortality*100000) 
```

Now, plot the mortality rate on the map.
```{r}
## convert state name to abbreviations for plotting

# Create a vector with state names and abbreviations
state_abbreviations = tibble(
  state_name = state.name, # Full state names
  state_abbr = state.abb  # Corresponding abbreviations
)

# Assuming covid_data is your dataset, join it with the state_abbreviations data
map_data = data_covid_df %>%
  left_join(state_abbreviations, by = c("state" = "state_name")) %>%  # Join by state name
  select(state, state_abbr, everything()) %>% 
  filter(!is.na(state_abbr)) 

```

```{r}
map_data_2021 = map_data %>% 
  filter(year == 2021)

# Create the choropleth map
fig = plot_ly(
  data = map_data_2021,
  type = "choropleth",
  locations = ~state_abbr, # State names
  locationmode = "USA-states", # Use state names in the USA
  z = ~covid_mortality_per100000, # Color scale based on covid_mortality
  text = ~paste("State: ", state), # Hover text
  colorscale = "Oranges",
  colorbar = list(title = "Mortality Rate per 100k")
) %>%
  layout(
    title = "COVID-19 Mortality by State (2021)", # Adjust the year in the title
    geo = list(
      scope = "usa", # Focus on the USA
      projection = list(type = "albers usa"), # Use Albers USA projection
      showlakes = TRUE,
      lakecolor = "rgb(255, 255, 255)" # Color of lakes
    )
  )

# Display the map
fig

```
let's try and add data of different years on this plot.

update: the plots for each year's data works well, but when combined together the colors are not showing, probably because of scale issues. Alternatively, we can make the plots for each year separately.
```{r}
# Assuming map_data is your dataset containing data from 2020-2023

# Create a choropleth map with dropdown to select years
fig = plot_ly(
  data = map_data,
  type = "choropleth",
  locations = ~state_abbr, # Use state abbreviations
  locationmode = "USA-states", # USA-state mode
  z = ~covid_mortality, # Color scale based on covid_mortality
  text = ~paste("State: ", state, "<br>Year: ", year, "<br>Mortality: ", round(covid_mortality, 5)), # Hover text
  colorscale = "Viridis",
  colorbar = list(
    title = "COVID Mortality",
    tickvals = seq(0, 0.002, length.out = 5),  # Set tickvals based on your range
    ticktext = seq(0, 0.002, length.out = 5)  # Adjust ticktext for clarity
  ),
  zmin = 0,  # Min value for the color scale
  zmax = 0.002  # Max value for the color scale
) %>%
  layout(
    title = "COVID-19 Mortality by State (2020-2023)",
    geo = list(
      scope = "usa",
      projection = list(type = "albers usa"),
      showlakes = TRUE,
      lakecolor = "rgb(255, 255, 255)"
    ),
    updatemenus = list(
      list(
        buttons = list(
          list(
            label = "2020",
            method = "relayout",
            args = list("geo.scope", "usa", "visible", TRUE, "z", map_data %>% filter(year == 2020) %>% pull(covid_mortality))
          ),
          list(
            label = "2021",
            method = "relayout",
            args = list("geo.scope", "usa", "visible", TRUE, "z", map_data %>% filter(year == 2021) %>% pull(covid_mortality))
          ),
          list(
            label = "2022",
            method = "relayout",
            args = list("geo.scope", "usa", "visible", TRUE, "z", map_data %>% filter(year == 2022) %>% pull(covid_mortality))
          ),
          list(
            label = "2023",
            method = "relayout",
            args = list("geo.scope", "usa", "visible", TRUE, "z", map_data %>% filter(year == 2023) %>% pull(covid_mortality))
          )
        ),
        direction = "down",
        showactive = TRUE,
        x = 1,
        xanchor = "left",
        y = 1.15,
        yanchor = "top"
      )
    )
  )

# Display the map
fig

```

```{r}
summary(map_data$covid_mortality)

# Check for the range of values
range(map_data$covid_mortality, na.rm = TRUE)

range(map_data_2021$covid_mortality, na.rm = TRUE)
```



