---
title: "analysis_jz"
author: "ELisajava"
date: "2024-11-18"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Load necessary libraries
```{r}
library(dplyr)
library(readr)
library(plotly)
```

# Import and tidy the data
```{r}
# Read the datasets
population_data <- read.table(file = "./data/Population by States.txt", header = TRUE, sep = "\t", quote = "\"", fill = TRUE)
death_data <- read_csv("./data/full_weekly_deaths_by_state_and_causes.csv")

# Clean and process population data
population_data_clean <- population_data %>%
  select(Residence.State, Year, Population) %>%
  filter(!is.na(Population)) %>%
  mutate(Year = as.numeric(gsub(" .*", "", Year))) %>%
  rename(State = Residence.State)

# Clean and process death data
death_data_clean <- death_data %>%
  select(state, year, all_cause, month) %>% 
  group_by(state, year, month) %>%
  summarise(Total_Deaths = sum(all_cause, na.rm = TRUE), .groups = "drop") %>%
  rename(State = state, Year = year, Month = month)

# Merge the datasets
merged_data <- death_data_clean %>%
  inner_join(population_data_clean, by = c("State", "Year"))

# Add state abbreviations
state_abbreviations <- data.frame(
  State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Florida",
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine",
            "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska",
            "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota",
            "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota",
            "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"),
  Abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS",
                   "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY",
                   "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV",
                   "WI", "WY")
)

# Merge abbreviations into the dataset
merged_data <- merged_data %>%
  left_join(state_abbreviations, by = "State") # Ensure 'Abbreviation' is added

# Define state regions
state_regions <- data.frame(
  Abbreviation = c("CT", "ME", "MA", "NH", "NJ", "NY", "PA", "RI", "VT", 
                   "IL", "IN", "IA", "KS", "MI", "MN", "MO", "NE", "ND", "OH", "SD", "WI", 
                   "AL", "AR", "DE", "DC", "FL", "GA", "KY", "LA", "MD", "MS", "NC", "OK", "SC", "TN", "TX", "VA", "WV", 
                   "AK", "AZ", "CA", "CO", "HI", "ID", "MT", "NV", "NM", "OR", "UT", "WA", "WY"),
  Region = c(rep("Northeast", 9), 
             rep("Midwest", 12), 
             rep("South", 17), 
             rep("West", 13))
)

# Merge regions into the dataset
merged_data <- merged_data %>%
  left_join(state_regions, by = "Abbreviation") # Join with region data

# Calculate mortality rates
merged_data <- merged_data %>%
  mutate(
    Annual_Mortality_Rate = (Total_Deaths / Population) * 1000,  # Annual rate per 1,000
    Monthly_Mortality_Rate = (Total_Deaths / Population) * 10000 / 12  # Monthly rate per 10,000
  )

# Save the updated dataset
write.csv(merged_data, "./data/Mortality_Rate_with_Region.csv", row.names = FALSE)

# Preview the result
head(merged_data)
```

# Monthly Mortality Rates in 2020 by Region and Stat
```{r}
# Filter for 2020 data
data_2020 <- merged_data %>%
  filter(Year == 2020)

# Create violin plots for each region with 2x2 layout and adjusted subheading positions
plot <- subplot(
  lapply(seq_along(unique(data_2020$Region)), function(i) {
    region_name <- unique(data_2020$Region)[i]
    region_data <- data_2020 %>% filter(Region == region_name)
    
    plot_ly(
      data = region_data,
      x = ~Abbreviation, 
      y = ~Monthly_Mortality_Rate,
      type = "violin",
      box = list(visible = TRUE),
      points = "all",
      color = ~Abbreviation, # Different colors for each state
      legendgroup = region_name,
      showlegend = FALSE
    ) %>%
      layout(
        annotations = list(
          list(
            text = paste0(region_name, " Region"), # Add region name as title
            x = 0.5, # Ensure each title is centered in its subplot
            y = 1.1, # Position above the subplot
            xref = "paper",
            yref = "paper",
            showarrow = FALSE,
            font = list(size = 16, color = "black"),
            xanchor = "center" # Center the text horizontally
          )
        ),
        xaxis = list(
          title = "State", # X-axis title
          titlefont = list(size = 12), # Reduce the size of axis title
          tickangle = 45 # Rotate the x-axis labels
        ),
        yaxis = list(
          title = "Monthly Mortality Rate (per 10,000)", # Y-axis title
          titlefont = list(size = 12) # Reduce the size of axis title
        )
      )
  }),
  nrows = 2, # Arrange in 2 rows
  shareY = TRUE, # Share the Y-axis
  titleX = TRUE, # Show X-axis titles
  titleY = TRUE, # Show Y-axis titles
  margin = 0.1 # Increase spacing between subplots
) %>%
  layout(
    title = list(
      text = "Monthly Mortality Rates in 2020 by Region and State", # Overall title
      x = 0.5, # Center the title
      y = 0.98, # Adjust vertical position
      font = list(size = 24) # Increase font size for the overall title
    ),
    margin = list(l = 80, r = 80, t = 120, b = 80), # Adjust global margins
    height = 800, # Increase the height of the overall plot
    width = 1000  # Increase the width of the overall plot
  )

# Render the final plot
plot
```
From the visualized data showing the **Monthly Mortality Rates (per 1,000 persons)** in 2020 and 2021 across different U.S. regions (South, West, Northeast, Midwest), the following observations can be made:

### 1. **Regional Variations**:
   - The **Northeast Region** generally shows higher peaks in monthly mortality rates compared to other regions, particularly noticeable in 2020. This could reflect the severe early impacts of COVID-19 in states like New Jersey, and Massachusetts. (Combined with the plots: COVID-19 mortality by state (2020-2023) ))
   - The **South Region** and **West Region** display relatively moderate mortality rates, with more stable patterns across states.
   - The **Midwest Region** shows some variability, with certain states like ND and SD having higher peaks than others.

### 2. **Temporal Trends**:
   - **2020** has more extreme variations in mortality rates compared to **2021**, especially in regions like the Northeast. This suggests that 2020 was more affected by large mortality spikes (likely due to the initial outbreak of COVID-19).
   - By **2021**, the monthly mortality rates appear more stable across most regions and states, which may indicate the effects of public health measures, vaccination campaigns, or improved healthcare responses.

### 3. **State-Level Observations**:
   - Certain states within regions exhibit outliers or higher peaks in mortality rates. For example:
     - **Northeast**: States like New Jersey and RI show high peaks in 2020.
     - **Midwest**: States like SD show noticeable variability.
     - **West**: Certain spikes are visible in states like MT, and NM, but overall patterns are less extreme.
     - **South**: States like Florida and Texas show slightly elevated peaks compared to others.

### 4. **Consistency Across Regions**:
   - The South and West regions show more consistent mortality rates across states, with fewer extreme variations compared to the Northeast and Midwest regions.

### 5. **Possible Influences**:
   - The observed variations could be influenced by factors such as:
     - COVID-19 outbreaks and associated mortality.
     - Healthcare capacity and access within regions and states.
     - Public health measures like lockdowns and mask mandates.

### Next Steps:
- **Explore Mortality Causes**: Break down the "all-cause mortality" into subcategories to determine whether the trends are dominated by COVID-19 or other factors like chronic diseases.
- **Compare by Population Density**: Normalize and compare across regions considering population density differences.
- **Policy Analysis**: Investigate how state or regional policies in response to the pandemic affected mortality rates.

# Monthly Mortality Rates in 2021 by Region and Stat
```{r}
# Filter for 2021 data
data_2021 <- merged_data %>%
  filter(Year == 2021)

# Define a custom color palette for states
state_colors <- c(
  "AL" = "#FF5733", "AK" = "#33FF57", "AZ" = "#3357FF", "AR" = "#FF33A1",
  "CA" = "#A133FF", "CO" = "#FFC300", "CT" = "#33FFF9", "DE" = "#FF5733",
  "FL" = "#C70039", "GA" = "#581845", "HI" = "#DAF7A6", "ID" = "#FFC300",
  "IL" = "#C70039", "IN" = "#900C3F", "IA" = "#FF5733", "KS" = "#33FF57",
  "KY" = "#FFC300", "LA" = "#581845", "ME" = "#DAF7A6", "MD" = "#FF33A1",
  "MA" = "#3357FF", "MI" = "#FF5733", "MN" = "#33FFF9", "MS" = "#FFC300",
  "MO" = "#C70039", "MT" = "#581845", "NE" = "#DAF7A6", "NV" = "#33FF57",
  "NH" = "#FFC300", "NJ" = "#FF5733", "NM" = "#33FFF9", "NY" = "#C70039",
  "NC" = "#FF33A1", "ND" = "#3357FF", "OH" = "#33FF57", "OK" = "#FF5733",
  "OR" = "#C70039", "PA" = "#FF33A1", "RI" = "#581845", "SC" = "#33FFF9",
  "SD" = "#FFC300", "TN" = "#FF5733", "TX" = "#33FF57", "UT" = "#FFC300",
  "VT" = "#FF33A1", "VA" = "#581845", "WA" = "#33FFF9", "WV" = "#C70039",
  "WI" = "#FF5733", "WY" = "#33FF57"
)

# Create violin plots for each region with 2x2 layout and adjusted subheading positions
plot <- subplot(
  lapply(seq_along(unique(data_2021$Region)), function(i) {
    region_name <- unique(data_2021$Region)[i]
    region_data <- data_2021 %>% filter(Region == region_name)
    
    plot_ly(
      data = region_data,
      x = ~Abbreviation, 
      y = ~Monthly_Mortality_Rate,
      type = "violin",
      box = list(visible = TRUE),
      points = "all",
      color = ~Abbreviation, # Different colors for each state
      colors = state_colors, # Use custom color palette
      legendgroup = region_name,
      showlegend = FALSE
    ) %>%
      layout(
        annotations = list(
          list(
            text = paste0(region_name, " Region"), # Add region name as title
            x = 0.5, # Ensure each title is centered in its subplot
            y = 1.1, # Position above the subplot
            xref = "paper",
            yref = "paper",
            showarrow = FALSE,
            font = list(size = 16, color = "black"),
            xanchor = "center" # Center the text horizontally
          )
        ),
        xaxis = list(
          title = "State", # X-axis title
          titlefont = list(size = 12), # Reduce the size of axis title
          tickangle = 45 # Rotate the x-axis labels
        ),
        yaxis = list(
          title = "Monthly Mortality Rate (per 10,000)", # Y-axis title
          titlefont = list(size = 12) # Reduce the size of axis title
        )
      )
  }),
  nrows = 2, # Arrange in 2 rows
  shareY = TRUE, # Share the Y-axis
  titleX = TRUE, # Show X-axis titles
  titleY = TRUE, # Show Y-axis titles
  margin = 0.1 # Increase spacing between subplots
) %>%
  layout(
    title = list(
      text = "Monthly Mortality Rates in 2021 by Region and State", # Overall title
      x = 0.5, # Center the title
      y = 0.98, # Adjust vertical position
      font = list(size = 24) # Increase font size for the overall title
    ),
    margin = list(l = 80, r = 80, t = 120, b = 80), # Adjust global margins
    height = 800, # Increase the height of the overall plot
    width = 1000  # Increase the width of the overall plot
  )

# Render the final plot
plot
```

# Monthly Mortality Rates in 2022 by Region and Stat
```{r}
# Filter for 2022 data
data_2022 <- merged_data %>%
  filter(Year == 2022)

# Create violin plots for each region with 2x2 layout and adjusted subheading positions
plot <- subplot(
  lapply(seq_along(unique(data_2022$Region)), function(i) {
    region_name <- unique(data_2022$Region)[i]
    region_data <- data_2022 %>% filter(Region == region_name)
    
    plot_ly(
      data = region_data,
      x = ~Abbreviation, 
      y = ~Monthly_Mortality_Rate,
      type = "violin",
      box = list(visible = TRUE),
      points = "all",
      color = ~Abbreviation, # Different colors for each state
      legendgroup = region_name,
      showlegend = FALSE
    ) %>%
      layout(
        annotations = list(
          list(
            text = paste0(region_name, " Region"), # Add region name as title
            x = 0.5, # Ensure each title is centered in its subplot
            y = 1.1, # Position above the subplot
            xref = "paper",
            yref = "paper",
            showarrow = FALSE,
            font = list(size = 16, color = "black"),
            xanchor = "center" # Center the text horizontally
          )
        ),
        xaxis = list(
          title = "State", # X-axis title
          titlefont = list(size = 12), # Reduce the size of axis title
          tickangle = 45 # Rotate the x-axis labels
        ),
        yaxis = list(
          title = "Monthly Mortality Rate (per 10,000)", # Y-axis title
          titlefont = list(size = 12) # Reduce the size of axis title
        )
      )
  }),
  nrows = 2, # Arrange in 2 rows
  shareY = TRUE, # Share the Y-axis
  titleX = TRUE, # Show X-axis titles
  titleY = TRUE, # Show Y-axis titles
  margin = 0.1 # Increase spacing between subplots
) %>%
  layout(
    title = list(
      text = "Monthly Mortality Rates in 2022 by Region and State", # Overall title
      x = 0.5, # Center the title
      y = 0.98, # Adjust vertical position
      font = list(size = 24) # Increase font size for the overall title
    ),
    margin = list(l = 80, r = 80, t = 120, b = 80), # Adjust global margins
    height = 800, # Increase the height of the overall plot
    width = 1000  # Increase the width of the overall plot
  )

# Render the final plot
plot
```

# Monthly Mortality Rates in 2023 by Region and Stat
```{r}
# Filter for 2023 data
data_2023 <- merged_data %>%
  filter(Year == 2023)

# Create violin plots for each region with 2x2 layout and adjusted subheading positions
plot <- subplot(
  lapply(seq_along(unique(data_2023$Region)), function(i) {
    region_name <- unique(data_2023$Region)[i]
    region_data <- data_2023 %>% filter(Region == region_name)
    
    plot_ly(
      data = region_data,
      x = ~Abbreviation, 
      y = ~Monthly_Mortality_Rate,
      type = "violin",
      box = list(visible = TRUE),
      points = "all",
      color = ~Abbreviation, # Different colors for each state
      legendgroup = region_name,
      showlegend = FALSE
    ) %>%
      layout(
        annotations = list(
          list(
            text = paste0(region_name, " Region"), # Add region name as title
            x = 0.5, # Ensure each title is centered in its subplot
            y = 1.1, # Position above the subplot
            xref = "paper",
            yref = "paper",
            showarrow = FALSE,
            font = list(size = 16, color = "black"),
            xanchor = "center" # Center the text horizontally
          )
        ),
        xaxis = list(
          title = "State", # X-axis title
          titlefont = list(size = 12), # Reduce the size of axis title
          tickangle = 45 # Rotate the x-axis labels
        ),
        yaxis = list(
          title = "Monthly Mortality Rate (per 10,000)", # Y-axis title
          titlefont = list(size = 12) # Reduce the size of axis title
        )
      )
  }),
  nrows = 2, # Arrange in 2 rows
  shareY = TRUE, # Share the Y-axis
  titleX = TRUE, # Show X-axis titles
  titleY = TRUE, # Show Y-axis titles
  margin = 0.1 # Increase spacing between subplots
) %>%
  layout(
    title = list(
      text = "Monthly Mortality Rates in 2023 by Region and State", # Overall title
      x = 0.5, # Center the title
      y = 0.98, # Adjust vertical position
      font = list(size = 24) # Increase font size for the overall title
    ),
    margin = list(l = 80, r = 80, t = 120, b = 80), # Adjust global margins
    height = 800, # Increase the height of the overall plot
    width = 1000  # Increase the width of the overall plot
  )

# Render the final plot
plot
```

# Generate data we need
```{r}
# Load necessary libraries
library(tidyr)
library(dplyr)
library(janitor)
library(readr)
library(ggplot2)

# Load the dataset using read_csv
data <- read_csv("./data/weekly_deaths_by_state_and_causes.csv")

# Convert column names to snake_case
data <- data %>%
  clean_names()

# Filter rows where jurisdiction_of_occurrence is "United States"
filtered_data <- data %>%
  filter(jurisdiction_of_occurrence == "United States")

# Display the filtered data
print(filtered_data)

# Count missing values for each column
missing_values <- filtered_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "column", values_to = "missing_values")

# Print all rows of the tibble
print(missing_values, n = nrow(missing_values))

# Remove columns from flag_allcause to flag_cov19ucod (inclusive)
filtered_data <- filtered_data %>%
  select(-flag_allcause:flag_cov19ucod)

# Display the modified dataset
print(filtered_data)

# Summarize data from all_cause to covid_19_u071_underlying_cause_of_death grouped by mmwr_year
summarized_data <- filtered_data %>%
  group_by(mmwr_year) %>%
  summarise(across(all_cause:covid_19_u071_underlying_cause_of_death, sum, na.rm = TRUE))

# Display the summarized data
print(summarized_data)

write.csv(summarized_data,"./data/summarized_data_united_states.csv")

# Read the data up to line 413
data <- read.table("./data/Population by States.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE, nrows = 413)


# Ensure column names are consistent
colnames(data) <- make.names(colnames(data))

# Aggregate the Population column by Year Code
summary_data <- aggregate(Population ~ Year.Code, data = data, FUN = sum)

# Save the aggregated data to a file
write.table(summary_data, "./data/Population_Summary_by_Year.txt", sep = "\t", row.names = FALSE, quote = FALSE)

print("Data has been aggregated and saved as Population_Summary_by_Year.txt")

# Read the datasets
population_data <- read.table("./data/Population_Summary_by_Year.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
summarized_data <- read_csv("./data/summarized_data_united_states.csv")

# Merge the datasets by matching mmwr_year and Year.Code
merged_data_usa <- merge(summarized_data, population_data, by.x = "mmwr_year", by.y = "Year.Code", all.x = TRUE)

# Get the column names for all death-related columns between septicemia_a40_a41 and covid_19_u071_underlying_cause_of_death
start_col <- "septicemia_a40_a41"
end_col <- "covid_19_u071_underlying_cause_of_death"

# Identify the range of columns between start_col and end_col inclusively
death_rate_columns <- merged_data_usa %>%
  select(all_of(start_col):all_of(end_col)) %>%
  colnames()

# Calculate death rates for each column in the range
for (col in death_rate_columns) {
  new_col_name <- paste0(col, "_death_rate")  # Generate new column name
  merged_data_usa[[new_col_name]] <- (merged_data_usa[[col]] / merged_data_usa$Population) * 100000
}

# View the first few rows to confirm the new columns
head(merged_data_usa)

# Rename columns in the dataset
merged_data_usa <- merged_data_usa %>%
  rename(
    septicemia = septicemia_a40_a41_death_rate,
    malignant_neoplasms = malignant_neoplasms_c00_c97_death_rate,
    diabetes = diabetes_mellitus_e10_e14_death_rate,
    alzheimer = alzheimer_disease_g30_death_rate,
    influ_pneum = influenza_and_pneumonia_j09_j18_death_rate,
    chronic_lower_respir = chronic_lower_respiratory_diseases_j40_j47_death_rate,
    other_respiratory = other_diseases_of_respiratory_system_j00_j06_j30_j39_j67_j70_j98_death_rate,
    nephritis_nephrotic = nephritis_nephrotic_syndrome_and_nephrosis_n00_n07_n17_n19_n25_n27_death_rate,
   not_classified = symptoms_signs_and_abnormal_clinical_and_laboratory_findings_not_elsewhere_classified_r00_r99_death_rate,
    diseases_of_heart = diseases_of_heart_i00_i09_i11_i13_i20_i51_death_rate,
    cerebrovascular = cerebrovascular_diseases_i60_i69_death_rate,
    covid_19_multiple_cause = covid_19_u071_multiple_cause_of_death_death_rate,
    covid_19_underlying_cause = covid_19_u071_underlying_cause_of_death_death_rate
  )

# Identify the columns to remove
columns_to_remove <- which(colnames(merged_data_usa) == "all_cause"):
                     which(colnames(merged_data_usa) == "Population")

# Remove the specified columns
filtered_data <- merged_data_usa[, -columns_to_remove]

# Duplicate each row
duplicated_data <- filtered_data[rep(1:nrow(filtered_data), each = 2), ]

# View the resulting dataset
head(duplicated_data)

# Optionally, save the resulting dataset to a new file
write.csv(duplicated_data, "./data/filtered_duplicated_data.csv", row.names = FALSE)

```

# Create Heatmap with mortality rate in each diseases in 2020
```{r}
# Load necessary libraries
library(plotly)
library(readr)

# Read the dataset
data <- read_csv("./data/filtered_duplicated_data.csv")  # Replace with the correct file path

# Filter the data for mmwr_year == 2020
data_2020 <- subset(data, mmwr_year == 2020)

# Remove the mmwr_year column to prepare for the heatmap
data_numeric <- data_2020[, -which(colnames(data_2020) == "mmwr_year")]

# Compute the correlation matrix (or use raw data values if correlation isn't needed)
correlation_matrix <- cor(data_numeric)

# Create the heatmap
fig <- plot_ly(
  x = colnames(correlation_matrix),  # Variables for x-axis
  y = colnames(correlation_matrix),  # Variables for y-axis (same as x-axis)
  z = correlation_matrix,            # Correlation or similarity values
  type = "heatmap",
  colors = colorRamp(c("white", "red"))  # Color scale
)

# Add layout details
fig <- fig %>% layout(
  title = "Heatmap for MMWR Year 2020",
  xaxis = list(title = "Variables"),
  yaxis = list(title = "Variables"),
  margin = list(l = 100, b = 100, t = 50, r = 50)
)

# Display the heatmap
fig
```
All grids other than the diagonal show NaN because cor() requires at least two distinct rows of data for each variable to compute correlations. In this case, the data for mmwr_year == 2020 may only have a single row, resulting in NaN values for all non-diagonal elements of the correlation matrix. Use Raw Values Instead of Correlations

# Create Raw Data Heatmap with mortality rate in each diseases in 2020
```{r}
# Load necessary libraries
library(plotly)
library(readr)

# Read the dataset
data <- read_csv("./data/filtered_duplicated_data.csv")  # Replace with the correct file path

# Filter the data for mmwr_year == 2020
data_2020 <- subset(data, mmwr_year == 2020)

# Remove the mmwr_year column to prepare for the heatmap
data_numeric <- data_2020[, -which(colnames(data_2020) == "mmwr_year")]

# Transpose the data so variables are both rows and columns
data_matrix <- t(as.matrix(data_numeric))

# Create the heatmap
fig <- plot_ly(
  x = colnames(data_matrix),  # Variable names for x-axis
  y = rownames(data_matrix),  # Variable names for y-axis
  z = data_matrix,            # Raw values
  type = "heatmap",
  colors = colorRamp(c("white", "red"))
)

# Add layout details
fig <- fig %>% layout(
  title = "Heatmap for MMWR Year 2020 (Raw Values)",
  xaxis = list(title = "Variables"),
  yaxis = list(title = "Variables"),
  margin = list(l = 100, b = 100, t = 50, r = 50)
)

# Display the heatmap
fig
```
